---
title: "Class09: Candy Analysis Mini Project"
author: "Yi-Hung Lee (PID:A16587141)"
format: pdf
---

Today we will examine the Halloween candy dataset

## 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

There are `r nrow(candy)` candy in the dataset

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## 2. What is your favorate candy?

```{r}
candy["Twix", ]$winpercent

## or

candy["Twix", "winpercent"]
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Kit Kat", "winpercent"]
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", "winpercent"]
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```


```{r}
skimr::skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The hist seems to be different

> Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
candy$chocolate
```

It means whether the specific candy contains chocolate

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
hist(candy$winpercent)

ggplot(candy, aes(winpercent)) +
  geom_histogram(bins = 20)
```

> Q9. Is the distribution of winpercent values symmetrical?

No

> Q10. Is the center of the distribution above or below 50%?

Less than 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
summary(candy$winpercent)
```

```{r}
chocolate_winpercent <- candy$winpercent[as.logical(candy$chocolate)]
fruity_winpercent <- candy$winpercent[as.logical(candy$fruity)]

mean(chocolate_winpercent)
mean(fruity_winpercent)
```

Chocolate is higher ranked than the fruity ones.

> Q12. Is this difference statistically significant?

```{r}
pVal <- t.test(chocolate_winpercent, fruity_winpercent)$p.value
```

It is `r pVal<0.05` to say this difference is statistically significant

## 3. Overall Candy Rankings

```{r}
head(candy[order(candy$winpercent), ])
```

> Q13. What are the five least liked candy types in this set?

```{r}
candy_order <- candy[order(candy$winpercent), ]
head(row.names(candy_order), n = 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(row.names(candy_order), n = 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy, aes(winpercent, rownames(candy))) +
  geom_col() 
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy, aes(winpercent, reorder(rownames(candy),winpercent))) +
  geom_col() 
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
#| fig-height: 13

ggplot(candy, aes(winpercent, reorder(rownames(candy),winpercent))) +
  geom_col(fill = my_cols)
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

## 4. Taking a look at pricepercent

```{r}
ggplot(candy, aes(winpercent, pricepercent, label = rownames(candy))) +
  geom_point(col = my_cols) +
  geom_label()
```

Too MANY LABELLLLSSS!!!!!

```{r}
library(ggrepel)

ggplot(candy, aes(winpercent, pricepercent, label = rownames(candy))) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols, size = 3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Super Bubble

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment()

```{r}
#| fig-height: 13
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col=my_cols) +
    geom_point()
```

## 5. Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and Fruity

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar

## 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col = my_cols, pch = 16)
```

```{r}
my_data <- cbind(candy, pca$x[, 1:3])

p <-  ggplot(my_data, aes(x = PC1, y = PC2, size = winpercent/100, label = rownames(my_data))) + 
  geom_point(col = my_cols) +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538") + 
  geom_text_repel(col = my_cols, size = 3, max.overlaps = 3)

p
```

```{r}


library(plotly)

# ggplotly(p)
```



How do the original variables contributes to our PCs? For this we look at the loading component of our results objects

```{r}
pca$rotation[,1]
```

```{r}
par(mar = c(8, 4, 4, 2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

```{r}
res <- pca$rotation

ggplot(res, aes(PC1, reorder(rownames(res), PC1))) + 
  geom_col() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ylab("Candy") 
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruit, Pluribus and hard are all picked up in the +ve direction. It make sense from the corralation from the candy experience of eating candy.
